# NPI Registry

```{r}
library(provider)
```

## Individual Providers

```{r}
nppes_npi(npi = 1710538772, 
          tidy = FALSE)
```


```{r}
atl_ind <- provider::nppes_npi(entype = "I", 
                               state = "GA", 
                               city = "Atlanta", 
                               tidy = FALSE)
atl_ind
```

Certification Date: NPPES allows providers to attest to the accuracy of their NPI data. When a provider requests any change to their NPI record, the certification date is updated. This may also effect the Last Updated column as well.


## Organizational Providers

```{r}
atl_org <- provider::nppes_npi(entype = "O", 
                               state = "GA", 
                               city = "Atlanta",
                               tidy = FALSE)
atl_org
```

```{r}
nppes_npi(1588375828)
```

Identifiers are used to associate other provider identifiers such as Medicaid or other insurers (ie:, Blue Cross, Blue Shield, Aetna, Kaiser-Permanente, etc.), with their NPI number. These identifiers can be used in matching an NPI number to an insurer's records. However, not all providers have such numbers and not all providers choose to include them in their NPI information.

<br>


```{r}
nucc_taxonomy_230
provider_enrollment(npi = 1417918293, specialty_code = "14-41")
```

```{r}
doctors_and_clinicians(enroll_id = "I20040309000221")
```

```{r}
facility_affiliations(
  npi = 1588375828
  #pac_id = 6103733050
  )

hospital_enrollment(city = "Atlanta")
```

```{r}
hospital_enrollment(city = "Atlanta") |> 
  dplyr::pull(npi) |> 
  purrr::map(nppes_npi) |> 
  purrr::list_rbind()
```

```{r}
nchar("I20040309000221")
nchar("O20041013000409")
```

